---
  title: "Multivariate"
author: "Aditya Shetty"
date: "3/19/2022"
output:
  word_document: default
html_document: default
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
rm(list=ls())
library(ggplot2)
library(cluster)
library(stats)
library(dplyr)
library(pls)
library(e1071)
```

## Introduction:

The data we are going to review is a part of dairy production for different breeds of cows and their characteristics which affect the quality of milk.  
This data which we are working on was curated for animal breeding purpose.  
As EU government gives more and more grants to the farmers, they have legal requirements to collect the data. Using the data collected government sanctions them subsidies.  
Due to this the amount of data collected is growing exponentially.  Ireland exports 90% of the dairy produce to 124 countries and is growing every year, due to which it has become essential and important for recording the data and analyzing.


## Objective we will cover in this project:

-> To understand the data and share our inference to it.  
-> Create clusters using unsupervised learning methods and compare them (hierarchical and k-means).  
-> Understanding PCA algorithm and implementing on our given data.  
-> Predicting value of alpha casin using the MIR spectra data.  
-> Creating a function which works similar to PLSR and predict.  

```{r}
# Loading the data set
milk<-read.csv("C:/Users/Aditya/Downloads/Milk_MIR_Traits_data.csv")
```

The data we will be using has dimension 431 rows of data and 582 columns.   It's a wide data.
  
The first 52 columns are the data durated by humans on different aspects eg: pH of the milk, fat content in milk, protein content in milk.  
MIR spectra are wavelengths calculated using the given data.  
There are 1060 wavelengths of MIR spectra and these data and columns are highly correlated and has multi collinearity.  
We have this data in columns 52 to 582.
  

  
```{r}
# Setting up the seed
set.seed(2100138)
# Generating random number between 1 to n (number of rows)
n<-sample(1:431,1)
# Deleting a row 
Deleted_milk<-milk[n,]
milk<-milk[-n,]
nrow(milk)  # checking number of rows after deleting one row
Deleted_milk
```


  

  
We can interpret from plot for alpha casein that it ranges from 5-40 and there are blank values in somwhere between 100th-200th observations.
  

```{r}
barplot(milk$alpha_s1_casein)
```
  
We can interpret from this plot that all MIR spectra data which are wavelengths ranges from 0-0.65 and they are highly collinear. 
  
```{r}
MIR <- as.matrix(milk[,52:582])
matplot(t(MIR), type='l', lty =1)
matplot(MIR)
```
  
We have 430 rows of data for different breeds.
  Once we remove all observation with alpha casein 3 standard deviations from mean, then 99.75% data is still stored in the mean range and outliers are removed.
  
Once we remove the observations we can observe that there is a difference in value as values above 24 in alpha casein are removed.  
  
  
```{r, warning=FALSE}
ggplot(milk, aes(x=milk$alpha_s1_casein)) +
  xlim(0,50) +
  ylim(0,90) +
  geom_histogram(color="blue", fill="white")
std <- sd(milk$alpha_s1_casein, na.rm = TRUE)
m <- mean(milk$alpha_s1_casein,na.rm = TRUE) 
lower <- m-3*std
higher <- m+3*std
data <- subset(milk, alpha_s1_casein <= higher & alpha_s1_casein >= lower)
#  2
ggplot(data, aes(x=data$alpha_s1_casein)) +
  xlim(0,50) +
  ylim(0,90) +
  geom_histogram(color="blue", fill="white")
```




  

    
Cluster Analysis is a term to describe technique used to achieve groups which are similar to each other.  
For this we use dissimilarity matrix which calculates the dissimilarity between each groups.  
  
Firstly, there are 3 major methods to calculate dissimilarity matrix:Euclidean method, Manhattan Method and Minkowski method. We  will go ahead with EUclidean method.
  
Then we have to select linkage between the groups:
  
There are 3 types of linkage, Single, Complete and Average linkage.
  
We cannot select linkage by number of clusters formed or the way clusters are formed. They can be selected by calculating coefficient of colligation which is a lengthy method. Both Average and complete are considered to be good approach, hence we will go with complete linkage for our analysis as there is very less limitations to it.
In single linkage there is a exception of chaining effect.

  


```{r}
milk_alpha<-data[,52:582]
############################
dist.eucl = dist(milk_alpha, method="euclidean")
E_complete = hclust(dist.eucl, method="complete")
plot(E_complete)
```
  
Above is the clusters we get based on euclidean complete Hierarchical clustering. 
  
K means method:  
  
```{r , echo=FALSE}
# Initializing Within group sum of squares
WGSS = rep(0,10)
n = nrow(data)
WGSS[1] = (n-1) * sum(apply(milk_alpha, 2, var))
for(k in 2:10)
{
  WGSS[k] = sum(kmeans(milk_alpha, centers = k)$withinss)
}
plot(1:10, WGSS, type="b", xlab="k", ylab="Within group sum of squares")
```
  
Here, Using elbow method we can conclude that we will have 6 or 7 optimal clusters. But since, it decreases very less after 6 to 7. We will conclude that k=6.  
  
```{r}
k=4
cl = kmeans(milk_alpha, center=k)
plot(scale(milk_alpha), col = cl$cluster)
points(cl$centers, col=1:k, pch=8, cex=2)
```
```{r}
hcl = cutree(E_complete, k=k)
pcl = kmeans(milk_alpha, centers=k)
tab = table(hcl, pcl$cluster)
tab
classAgreement(tab)
```
  
As we can see using Rand index we get above 74% match, which is good when we compare hierarchical cluster and K means cluster.


  
### (PCA)

```{r}
pca=prcomp(scale(milk_alpha))
plot(pca, main="Cumulative proportion of the variance" ,xlab="PCA1 -PCA10", xaxt='n')
axis(1, at = seq(1,11, by = 1.1),labels = 1:10)
# summary(pca)
```
  
As we notice in proportion of Variance the first 4 columns contain (63.70% + 18.76% + 12.69% + 4.078%) 99.228% of the variance.
  
Also with the screeplot we can confirm that we can reduce the dimension of columns to 4 for Principal Component scores. i.e We can use the first 4 columns to retain 99.228% information and reduce the redundancy in the given data.
  


##  5
  

```{r}
#Getting the covariance matrix sigma from data
S=cov(milk_alpha)
#Getting eigen values from covariance matrix E
eigen=eigen(S)
#Getting eigen vectors
eigenvectors=eigen$vectors
#Getting PCS value by data*eigenvectors 
scores<- (as.matrix(scale(milk_alpha, center = TRUE))) %*%  (eigenvectors)
#Using predict function just to compare with our data
pred<-predict(pca)
par(mfrow=c(1,2))
barplot(scores, main="PCS scores using eigenvectors from data")
barplot(pred, main="PCS scores using pred function")
par(mfrow=c(2,1))
plot(scores[,1], scores[,2], xlab="PC1", ylab="PC2", main = "USING eigenvectors from data to get PCS value")
plot(pred[,1], pred[,2], xlab="PC1", ylab="PC2", main = "USING In built pred function for PCS")
```

  
As we can see what we achieved is close to what we achieve from pred function in PCA.
pca$rotation basically returns eigen vectors, if we notice there is a change in sign due to which we receive mirror image of our data which we get using predict, but the essence and variation remains the same. This is because eigenvectors in different compilers gives different signs. Hence, the mirror image (referring help page from R: {r} [https://stat.ethz.ch/pipermail/r-help/2003-July/036504.html] (https://stat.ethz.ch/pipermail/r-help/2003-July/036504.html) ). 

  

We will get exact same figures if we use pca$rotation which is also the eigenvectors of our data.
  
```{r}
scores<- (as.matrix(scale(milk_alpha, center = TRUE))) %*%  (pca$rotation)
par(mfrow=c(1,2))
barplot(scores, main="PCS scores using rotation from prcomp")
barplot(pred, main="PCS scores using pred function")
par(mfrow=c(2,1))
plot(scores[,1], scores[,2], xlab="PC1", ylab="PC2", main = "USING rotation from prcomp to get PCS value")
plot(pred[,1], pred[,2], xlab="PC1", ylab="PC2", main = "USING In built pred function for PCS")
```

  
Principal component Regression:
  
Purpose: The main purpose of it is dimension reduction, Retain majority of the variance and importance while doing so. It is unsupervised way because we are only using the explanatory variable and not target variables for prediction while training/fitting model. PCR is used in order to prevent errors caused by dependencies between assumed independent variables in regression.
  
Description on how method works:   
-> First we need to scale and center the data .i.e normalize the data.  
-> We will divide the dataframe into 2 parts taking one third as test data and two third as train data.  
-> Run PCA from prcomp.
-> Select number of principal components which has the majority using proportion of variance.  
-> Calculate the Principal Component Scores using the principal components/columns.
-> Run PCR function to get target variable response and use K-fold method for cross validation of response.
Choices we need to make: We need scale and center the data. Selecting number of principal components. If we need to k-fold method.
  
Advantages/Disadvantages: Multi-Collinearity disappears since we reduce the dimension and only take ones with majority variance. Numerical accuracy is improved due to the use of Principal Components. It has less accuracy than PLSR since the data is modelled independently without taking target variable in consideration.
  
Partial Least Square Regression:
  
Purpose:The main purpose of it is dimension reduction, Retain majority of the variance and importance while doing so. It doesn't use Principal component but uses latent variables. PLSR is uses target variable along with explanatory values in order to determine the coefficients and intercept value, hence we get better accuracy than PCR.

Description on how method works:   
  -> We will divide the dataframe into 2 parts taking one third as test data and two third as train data.  
-> Run PLSR function from pls package with scaling and validation as TRUE.
-> Select number of latent variables using RMSEP plots where till CV reduces or majoirt variance is covered.  
-> Calculate the Principal Component Scores using the principal components/columns.
-> Run PCR function to get target variable response and use K-fold method for cross validation of response.
Choices we need to make: We need scale and center the data. Selecting number of latent variables based on proportion of variance. If we need k-fold method to cross validate.

Advantages/Disadvantages: Multi-Collinearity disappears since we reduce the dimension and only take ones with majority variance. Better accuracy than PCR because it uses the target variable for training. Disadvantage of PLSR is that it is more prone to overfitting. 





```{r}
alpha_s1_casein = data[,9:9]
n=nrow(milk_alpha)
dataPCR <- cbind(alpha_s1_casein,data[,52:582]) 
dataPCR_train <- dataPCR[1:((2*n/3)-1),]
dataPCR_test <- dataPCR[(2*n/3):n,]
#Model fitting: Training model
plsrs <- plsr(alpha_s1_casein ~.,ncomp=10, data = dataPCR_train , scale = TRUE , validation = "CV")
plot(RMSEP(plsrs), legendpos = "topright")
```

We will take number of components as 6, because CV-RMSEP reduces till 6 and then starts increasing after that, hence number of latent component selected is 6.

```{r}
#Component number 
plsrs.RMSEP = RMSEP(plsrs, estimate="CV")
min_comp = which.min(plsrs.RMSEP$val)-1
#We have it
min_comp
plot(plsrs, ncomp=min_comp, asp=1, line=TRUE)
pls.pred = predict(plsrs, dataPCR_test, ncomp=min_comp)
#Mean Squared error for our prediction
sqrt(mean((pls.pred - dataPCR_test$alpha_s1_casein)^2))
```




```{r}
#I am running first iteration outside to create data frame for T,W,Q,P and appending them.
#From second to 6th iteration I am running it in loop.
#Training the model to get coefficient
X <- dataPCR_train[,2:532]
Y <- dataPCR_train[,1:1]
# SVD (x^t * Y)
S <- t(as.matrix(X)) %*% as.matrix(Y)
SVD <- svd(S)
#First vector from U and V respectively, but both have one columns only
w <- SVD$u
q <- SVD$v
t <- as.matrix(X) %*% as.matrix(w)
u <- as.matrix(Y) %*% as.matrix(q)
abs_t <- t(as.matrix(t)) %*% as.matrix(t)
t <- t/(sqrt(abs_t[1,1]))
En <- X
Fn <- Y
p <- t(as.matrix(En))%*% t
q <- t(as.matrix(Fn))%*% t
En <- En - t %*% t(p)
Fn <- Fn - t %*% t(q)
T <- t
W <- w
Q <- q
P <- p
#Second iteration to sixth component taking previous  as reference
for (i in 2:6) {
  X <- En
  Y <- Fn
  S <- t(as.matrix(X)) %*% as.matrix(Y)
  SVD <- svd(S)
  w <- SVD$u
  q <- SVD$v
  t <- as.matrix(X) %*% as.matrix(w)
  u <- as.matrix(Y) %*% as.matrix(q)
  abs_t <- t(as.matrix(t)) %*% as.matrix(t)
  t <- as.matrix(t) / (sqrt(abs_t[1,1]))
  En <- X
  Fn <- Y
  p <- t(as.matrix(En))%*% t
  q <- t(as.matrix(Fn))%*% t
  En <- En - as.matrix(t) %*% t(p)
  Fn <- Fn - as.matrix(t) %*% t(q)
  T <- cbind(T,t)
  W <- cbind(W,w)
  Q <- cbind(Q,q)
  P <- cbind(P,p)
  
}
R= as.matrix(W) %*% solve(t(as.matrix(P)) %*% as.matrix(W))
#Calculated the required Coefficients
B= R %*% t(Q)
Pred_Y = as.matrix(X) %*% as.matrix(B)
#Mean square error for showing accuracy
sqrt(mean((Pred_Y - Y)^2)) 
summary(Pred_Y)
summary(Y)
```

```{r}
#Prediction for test data
Predicted_test_alpha = as.matrix(dataPCR_test[,2:532]) %*% B 
sqrt(mean((Predicted_test_alpha - dataPCR_test[,1:1])^2))
plot(Predicted_test_alpha, type="l",lwd=3, ) #Predicted value
points(dataPCR_test[,1:1], col='red') #Actual data 
grid (10,10, lty = 6, col = "cornsilk2")
summary(Predicted_test_alpha)
summary(dataPCR_test[,1:1])
```
As we can see the minimum and maximum values are in range and they match.  
In the plot we can see that the lines match and cover most of the actual data, which shows pur prediction was close.
